---
title: "ALL TranscriptomicsQC"
author: "Shubham Agrawal"
date: "2025-07-14"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png","pdf"), fig.path = "./ALLTranscriptomicsQC/")
library(tidyverse)
library(DESeq2)
library(ggrepel)
source("helper.R")
```

# Upload the object

```{r, message=FALSE}
dds <- readRDS("/Users/shubhamagrawal/Desktop/ALL/dds.Rds")
dds
```


# Pre-Filtering

Objectives:
- by removing rows in which there are very few reads, we reduce the memory size of the dds data object, and we increase the speed of count modeling within DESeq2
- it can also improve visualizations, as features with no information for differential expression are not plotted in dispersion plots or MA-plots.

Here we perform pre-filtering to keep only rows that have a count of at least 10 for a minimal number of samples.

```{r}
dim(dds)
# remove genes never expressed
#dds <- dds[rowSums(counts(dds)) != 0,]
smallestGroupSize <- 100
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep, ]
dim(dds)
```

```{r}
plotBoxInstensities(se = dds)
```

# Perform vst

```{r}
ddsVst <- varianceStabilizingTransformation(dds)
```

```{r}
plotBoxInstensities(se = ddsVst)
```

```{r}
plotHeatDDS(dds = ddsVst)
```

```{r}
pcaData <- DESeq2::plotPCA(ddsVst, intgroup=c("diseaseType", "diagnosisType", "Duplicate_RNA_Sample_Available."), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=diseaseType, shape=diagnosisType)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()

ggplot(pcaData, aes(PC1, PC2, color=diseaseType, shape=diagnosisType, label=Duplicate_RNA_Sample_Available.)) +
  geom_point(size=3) +
  #geom_text_repel(aes(label = ifelse(expression %in% c("Up", "Down"), as.character(Gene),"")), max.overlaps = Inf) +
  geom_label() +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

Points to note from the PCA and heatmap:
1. PCA group T-ALL and B-ALL into distinct clusters. Good.
2. Sample swapping. Some T-ALL samples are in the B-ALL cluster and same for 
some B-All samples
4. One sample is contaminated: OE0583_T-ALL_5892
3. Replicate samples are next to each other. Remove the replicates.

# Removing replicate and contaminated samples

```{r}
contaminated <- "OE0583_T-ALL_5892"
dds <- dds[, dds$temp != contaminated]
dds
```

```{r}
# Subset dds to keep only replicate samples
ddsReplicate <- dds[, !is.na(dds$Duplicate_RNA_Sample_Available.)]

# Get sample IDs from colData
sample_ids <- ddsReplicate$Duplicate_RNA_Sample_Available.

# Get total counts per sample (library size)
total_counts <- colSums(counts(ddsReplicate))

# Combine into a data frame
df_counts <- data.frame(
  sampleID = sample_ids,
  totalCounts = total_counts,
  colname = colnames(ddsReplicate)
)

# For each sampleID, keep the column name with lowest total counts
removeSamples <- df_counts %>%
  group_by(sampleID) %>%
  slice_min(order_by = totalCounts, n = 1, with_ties = FALSE) %>%
  pull(colname)

# remove the replicates with the lower counts
ddsFiltered <- dds[, !colnames(dds) %in% removeSamples]
ddsFiltered
```
```{r}
saveRDS(ddsFiltered, file = "ddsFiltered.Rds")
```


# Check the plots again

```{r}
ddsFilteredVst <- varianceStabilizingTransformation(ddsFiltered)
```

```{r}
plotBoxInstensities(se = ddsFilteredVst)
```

```{r}
plotHeatDDS(dds = ddsFilteredVst)
```
```{r}
pcaDataFiltered <- DESeq2::plotPCA(ddsFilteredVst, intgroup=c("diseaseType", "diagnosisType", "Duplicate_RNA_Sample_Available."), returnData=TRUE)
percentVar <- round(100 * attr(pcaDataFiltered, "percentVar"))
ggplot(pcaDataFiltered, aes(PC1, PC2, color=diseaseType, shape=diagnosisType)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()

ggplot(pcaDataFiltered, aes(PC1, PC2, color=diseaseType, shape=diagnosisType, label=Duplicate_RNA_Sample_Available.)) +
  geom_point(size=3) +
  #geom_text_repel(aes(label = ifelse(expression %in% c("Up", "Down"), as.character(Gene),"")), max.overlaps = Inf) +
  geom_label() +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

Check different CD markers to find if the labelling is correct or not

B-ALL: CD19, CD20, CD10, CD22, CD79A, CD34, CD38
T-ALL: CD2, CD3, CD5, CD7, CD8 or CD4, CD1A, CD44, CD99

```{r}
# Suppose your DESeq2 dataset is named `dds`
# Get the count matrix (raw or normalized)
ddsFiltered1 <- ddsFiltered[, !is.na(ddsFiltered$diseaseType)]
counts_mat <- counts(ddsFiltered1)
counts_mat <- log2(counts_mat + 1)

# Extract gene annotation (rowData)
gene_anno <- as.data.frame(rowData(ddsFiltered1))

# Example: list of CD markers of interest
cd_genes <- c("CD19", "CD20", "CD22", "CD10", "CD3", "CD7", "CD5", "CD2", "CD4", "CD8A", "CD8B", "CD1A")

# Find rows corresponding to these gene names
selected_genes <- gene_anno %>%
  filter(name %in% cd_genes)

# selected_genes <- gene_anno %>%
#   filter(grepl("^CD[0-9]+", name, ignore.case = TRUE))

# Subset the normalized counts matrix for these genes
cd_counts <- counts_mat[rownames(counts_mat) %in% selected_genes$gene_id, ]

# (Optional) rename rows to gene names for clarity
rownames(cd_counts) <- selected_genes$name[
  match(rownames(cd_counts), selected_genes$gene_id)]

# Extract disease type information for annotation
sample_anno <- as.data.frame(colData(ddsFiltered1)) %>%
  select(diseaseType)

# # Create the heatmap
# pheatmap(cd_counts,
#          scale = "row",                  # z-score scaling across genes
#          annotation_col = sample_anno,   # adds diseaseType annotation
#          show_rownames = TRUE,
#          cluster_cols = TRUE,
#          cluster_rows = TRUE,
#          fontsize_row = 10,
#          fontsize_col = 9,
#          main = "CD Marker Expression Heatmap")


pheatmap(cd_counts,
         scale = "row",
         annotation_col = sample_anno,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         main = "CD Marker Expression by Disease Type")



```


# Association tests for the PCs

```{r}
pca <- stats::prcomp(t(assay(ddsFilteredVst)), center = TRUE, scale. = TRUE)
resPCA <-  as.data.frame(pca$x)
resPCA <- resPCA[ , 1:25]
resPCA$samples <- rownames(pca$x)
cd <- as.data.frame(colData(ddsFilteredVst))
cd <- subset(cd, select = -c(Duplicate_RNA_Sample_Available., 
                             Protein_Sample_ID, temp, patID))

resAssc <- testAssociation(tabX = resPCA, tabY = cd, joinID = "samples", 
                           correlation_method = "pearson", 
                           plot = TRUE, pCut = 0.01, ifFdr = FALSE, pMax = 12, 
                           onlySignificant = FALSE)

p <- resAssc$plot
p
```

# Subsetting the transcriptomics samples based on disease

```{r}
ddsFilteredVst <- ddsFilteredVst[, !is.na(ddsFilteredVst$diseaseType) &
                                   !is.na(ddsFilteredVst$diagnosisType)]
print("Filtered Vst B-ALL")
ddsB <- ddsFilteredVst[, ddsFilteredVst$diseaseType == "B-ALL"]
ddsB
print("Filtered Vst T-ALL")
ddsT <- ddsFilteredVst[, ddsFilteredVst$diseaseType == "T-ALL"]
ddsT
```
```{r}
pcaB <- DESeq2::plotPCA(ddsB, intgroup=c("diagnosisType", "Duplicate_RNA_Sample_Available."), returnData=TRUE)
percentVar <- round(100 * attr(pcaB, "percentVar"))
ggplot(pcaB, aes(PC1, PC2, color=diagnosisType)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  ggtitle("PCA for transcriptomics B-ALL data")
```
```{r}
pcaT <- DESeq2::plotPCA(ddsT, intgroup=c("diagnosisType", "Duplicate_RNA_Sample_Available."), returnData=TRUE)
percentVar <- round(100 * attr(pcaT, "percentVar"))
ggplot(pcaT, aes(PC1, PC2, color=diagnosisType)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  ggtitle("PCA for transcriptomics T-ALL data")
```

```{r}
pca <- stats::prcomp(t(assay(ddsB)), center = TRUE, scale. = TRUE)
resPCA <-  as.data.frame(pca$x)
resPCA <- resPCA[ , 1:25]
resPCA$samples <- rownames(pca$x)
cd <- as.data.frame(colData(ddsB))
cd <- subset(cd, select = -c(Duplicate_RNA_Sample_Available., 
                             Protein_Sample_ID, temp, patID))

resAssc <- testAssociation(tabX = resPCA, tabY = cd, joinID = "samples", 
                           correlation_method = "pearson", 
                           plot = TRUE, pCut = 0.01, ifFdr = FALSE, pMax = 12, 
                           onlySignificant = FALSE)

p <- resAssc$plot
p
```

```{r}
pca <- stats::prcomp(t(assay(ddsT)), center = TRUE, scale. = TRUE)
resPCA <-  as.data.frame(pca$x)
resPCA <- resPCA[ , 1:25]
resPCA$samples <- rownames(pca$x)
cd <- as.data.frame(colData(ddsT))
cd <- subset(cd, select = -c(Duplicate_RNA_Sample_Available., 
                             Protein_Sample_ID, temp, patID))

resAssc <- testAssociation(tabX = resPCA, tabY = cd, joinID = "samples", 
                           correlation_method = "pearson", 
                           plot = TRUE, pCut = 0.01, ifFdr = FALSE, pMax = 12, 
                           onlySignificant = FALSE)

p <- resAssc$plot
p
```

```{r}
ddsFilteredVst
```

