---
title: "Drug Response Prediction"
author: "Shubham Agrawal"
date: "2026-01-28"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(tidyverse)
library(DESeq2)
library(glmnet)
library(RColorBrewer)
library(reshape2)
```

# Load the data

```{r}
seProt <- readRDS("protoemics.Rds")
dds <- readRDS("ddsFiltered.Rds")
seDRP <- readRDS("drpObject.Rds")
```

# Get only T-ALL

```{r}
seProt <- seProt[, seProt$diseaseType == "T-ALL"]
print("Proteomics T-ALL data")
seProt
dds <- dds[, !is.na(dds$diseaseType) & !is.na(dds$diagnosisType)]
dds <- dds[, dds$diseaseType == "T-ALL"]
print("Transcriptomics T-ALL data")
dds
seDRP <- seDRP[, seDRP$diseaseType == "T-ALL"]
print("Drug Response Profile T-ALL data")
seDRP
```

# Create unique identifier

Get the colData from dds and drp objects
```{r}
cdProt <- as.data.frame(colData(seProt))
#head(cdProt)
cdDDS <- as.data.frame(colData(dds))
#head(cdDDS)
cdDDS <- select(cdDDS, samples, patID, Protein_Sample_ID)
cdDRP <- as.data.frame(colData(seDRP))
#head(cdDRP)
cdDRP <- select(cdDRP, Labeling.proteomics, sample_id)

cdjoin <- left_join(cdProt, cdDDS, by = c("sample" = "Protein_Sample_ID"))
cdjoin <- left_join(cdjoin, cdDRP, by = c("sample" = "Labeling.proteomics"))
cdjoin <- cdjoin[!duplicated(cdjoin$sample), ]
```
 S24, S33, S63, S74
 
```{r}
cdFinal <- cdjoin[!duplicated(cdjoin$sample), ]
dim(cdFinal)
names(cdFinal)[names(cdFinal) == "sample"] <- "protSample"
names(cdFinal)[names(cdFinal) == "samples"] <- "rnaSample"
names(cdFinal)[names(cdFinal) == "sample_id"] <- "drugSample"
```

```{r}
if (!"sample" %in% colnames(cdFinal)) cdFinal$sample <- as.character(seq_len(nrow(cdFinal)))
rownames(cdFinal) <- cdFinal$sample
commonSamples <- cdFinal$sample
```

# Normalize RNAseq data

```{r}
minrs <- 100
rs  <- rowSums(assay(dds))
dds <- dds[rs >= minrs, ]
dds
# variance stabilize the data
# (includes normalizing for library size and dispsersion estimation) 
vsd <- varianceStabilizingTransformation(dds)
# no NAs in data
any(is.na(assay(vsd)))
```
```{r}
# helper to align a matrix to common_samples given the sample_map mapping
alignView <- function(mat, from_ids, mapping_df, common_ids, colname_in_map) {
  # mat: original matrix (rows = features, cols = samples, colnames = sample IDs used in mat)
  # from_ids: character vector of sample IDs in mat (should equal colnames(mat))
  # mapping_df: the sample_map data.frame
  # common_ids: mapping_df$commonID
  # colname_in_map: column name in mapping_df that links to this view (e.g. "prot_sample")
  out <- matrix(NA_real_, nrow = nrow(mat), ncol = length(common_ids),
                dimnames = list(rownames(mat), as.character(common_ids)))
  # for each row in mapping, if sample exists in mat, copy column
  for (i in seq_along(common_ids)) {
    mapped_name <- mapping_df[[colname_in_map]][i]
    if (!is.na(mapped_name) && mapped_name %in% colnames(mat)) {
      out[, i] <- mat[, mapped_name]
    } else {
      # leave NA column
    }
  }
  return(out)
}

# Proteomics view (assay might already be normalized, imputed)
protMat <- as.matrix(assay(seProt))  
# ensure colnames are the sample names used in sample_map$prot_sample
protView <- alignView(protMat, colnames(protMat), cdFinal, commonSamples, "protSample")

# Drug view (logAUC)
drugMat <- as.matrix(assay(seDRP))      # rows = drugs, cols = sampleIDs (drug sampleIDs)
drugView <- alignView(drugMat, colnames(drugMat), cdFinal, commonSamples, "drugSample")

# RNA counts view: use normalized counts assay from DESeq2 object
countsMat <- as.matrix(assay(vsd))
rnaView <- alignView(countsMat, colnames(countsMat), cdFinal, commonSamples, "rnaSample")

# Optional: sanity check dims and names
dim(protView); dim(rnaView); dim(drugView)
colnames(protView)  # should be common_samples (or same vector across views)
```

# Feature selection and filtering

```{r}
selectTopVar <- function(mat, topN = 2000) {
  varVec <- matrixStats::rowVars(mat, na.rm = TRUE)
  keep <- order(varVec, decreasing = TRUE)[seq_len(min(topN, sum(!is.na(varVec))))]
  mat[keep, , drop = FALSE]
}

rnaViewFilt <- selectTopVar(rnaView, topN = 3000)     
protViewFilt <- selectTopVar(protView, topN = 2000)
drugViewFilt <- drugView   # usually small number of drugs
```

# Identify valid samples

```{r}
valid_samples <- Reduce(intersect, list(
  colnames(rnaViewFilt)[colSums(!is.na(rnaViewFilt)) > 0],
  colnames(protViewFilt)[colSums(!is.na(protViewFilt)) > 0],
  colnames(drugViewFilt)[colSums(!is.na(drugViewFilt)) > 0],
  rownames(cdFinal)
))
```

Subset all data:

```{r}
rna <- rnaViewFilt[, valid_samples]
prot <- protViewFilt[, valid_samples]
drug <- drugViewFilt[, valid_samples]
clin <- cdFinal[valid_samples, ]
```

# Extract Sex and Age and diagnosisType from clinical data

```{r}
clin$Sex[is.na(clin$Sex)] <- "Unknown"
clin$Sex <- factor(clin$Sex, levels = c("Female", "Male", "Unknown"))
sex_mat <- model.matrix(~ Sex - 1, data = clin)
```

```{r}
clin$Age <- as.numeric(clin$Age)
clin$Age[is.na(clin$Age)] <- median(clin$Age, na.rm = TRUE)
age_scaled <- scale(clin$Age)
colnames(age_scaled) <- "Age"
```

```{r}

clin$diagnosisType <- factor(clin$diagnosisType, levels = c("Primary", "Relapse"))
diagnosis_mat <- model.matrix(~ diagnosisType - 1, data = clin)
```

```{r}
clinMat <- cbind(
  Age = age_scaled,
  Sex = sex_mat,
  Diagnosis = diagnosis_mat
)
```

# Use top 20 PCs of expression and proteomics data

```{r}
pcaExpr<-prcomp(t(rna), center=T, scale. = F)
rnaPCs<-pcaExpr$x[,1:20]
colnames(rnaPCs) <- paste("rna",colnames(rnaPCs), sep="")
```

```{r}
pcaProt<-prcomp(t(prot), center=T, scale. = F)
protPCs<-pcaProt$x[,1:20]
colnames(protPCs) <- paste("prot",colnames(protPCs), sep="")
```

# Choose ONE Drug (Response Variable)

```{r}
# drug_name <- rownames(drug)[10]   # choose drug
# y <- as.numeric(drug[drug_name, ])
# names(y) <- colnames(drug)
```

```{r}
# Remove samples with missing Y:
# keep <- !is.na(y)
# y <- y[keep]
# rna <- rnaPCs[keep, ]
# prot <- protPCs[keep, ]
# clin <- clinMat[keep, ]
```

```{r}
ZPCs<-list(transcriptomics=rnaPCs,
        proteomics=protPCs, 
        demographics=clinMat)
ZPCs$all<-do.call(cbind, ZPCs)
ZPCsunscaled<-ZPCs
ZPCsscaled<-lapply(ZPCs, scale)
lapply(ZPCsscaled, colnames)
```
Define colors.

```{r}
set1 <- brewer.pal(9,"Set1")
colMod<-c(paste(set1[c(1,5,2)],"88",sep=""), "grey")
names(colMod) <-
  c("demographics", "transcriptomics", "proteomics", "all")
```

```{r}
#Function to calculate Var Explained for Penalized Regression
R2ForPenRegPlusViz <- function(Z, drugviab, nfolds=10, alpha=1, nrep=100,
                      Parmfrow=c(2,4), ylimMax=0.4, standardize=TRUE) {
  Zlocal<-Z

  set.seed(1030)
  seeds<-sample(1:10000000, nrep)

  RepeatedR2list<-lapply(1:nrep, function(outer){
  # Use same folds for all omics to make comparable
  set.seed(seeds[outer])
  foldsAssignment<-sample(rep(seq(nfolds), length=ncol(drugviab)))

  R2echOmicadj<-sapply(rownames(drugviab), function(dname) {
    
    # filter out samples where drug response is missing
    d <- as.numeric(drugviab[dname, ])
    names(d) <- colnames(drugviab)
    keep <- !is.na(d)
    d <- d[keep]
    Zlocal$transcriptomics <- Zlocal$transcriptomics[keep, ]
    Zlocal$proteomics <- Zlocal$proteomics[keep, ]
    Zlocal$demographics <- Zlocal$demographics[keep, ]
    Zlocal$all <- Zlocal$all[keep, ]
    
    
    
    sapply(names(Zlocal), function(nameZ) {
      pred<-Zlocal[[nameZ]]
      #fit a lasso model for omics with more than one features
      if(ncol(pred)>1){
      fitcv<-cv.glmnet(pred,d,family="gaussian", standardize=standardize,
                        alpha=alpha)
      R2 <- 1-min(fitcv$cvm)/fitcv$cvm[1]
      
      
      }
      else {#fit a liner model for single feautres (IGHV)
        fitlm<-lm(d~., data.frame(pred))
        R2 <- summary(fitlm)$r.squared
      }
      R2
    })
  }
    )
})
RepeatedR2<-RepeatedR2list
#calculate mean and sd across repetitions with different folds
meanR2<-apply(simplify2array(RepeatedR2), 1:2, mean)
sdR2<-apply(simplify2array(RepeatedR2), 1:2, sd)

par(mfrow=Parmfrow, mar=c(7,5,3,3))
for (i in 1: ncol(meanR2)) {
    barc<-barplot(meanR2[,i], main= colnames(meanR2)[i], ylim=c(0,ylimMax),
                  las=2, col=colMod[rownames(meanR2)], ylab="R2")
    segments(barc, meanR2[,i]-sdR2[,i],barc, meanR2[,i]+sdR2[,i])
}
RepeatedR2
}
```

Fit model and show resulting omic-prediction profiles.
```{r}
resultLassoPCA<-R2ForPenRegPlusViz(Z = ZPCsscaled, drugviab = drug, nfold=10, alpha=1,
                                   nrep=100, ylimMax=0.4)

```

```{r, fig.path="./figureDRP/", dev=c("png", "pdf"), fig.width=9, fig.height=8}}
df_resultLassoPCA <- melt(resultLassoPCA)
colnames(df_resultLassoPCA) <- c("omic", "drug", "R2", "run")
summaryR2 <- df_resultLassoPCA %>% group_by(omic, drug) %>% 
  dplyr::summarise(meanR2=mean(R2),sdR2 = sd(R2), nR2 = length(R2)) %>%
  mutate(seR2 = sdR2/sqrt(nR2))

ggplot(summaryR2, aes(x=omic, y=meanR2, fill=omic, group=omic))+
    geom_bar(stat = "identity") +  scale_fill_manual(values=colMod) + 
    geom_errorbar(aes(ymax = meanR2 + sdR2,ymin = meanR2 - sdR2), position = "dodge", width = 0.25) +facet_wrap(~drug, ncol=4) +theme_bw(base_size = 18) +ylab(bquote(R^2)) + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.text = element_text(colour = "black")) +xlab("")+guides(fill=guide_legend(title="Data type")) 
```

```{r}
resultLassoPCA5 <-R2ForPenRegPlusViz(Z = ZPCsscaled, drugviab = drug, nfolds=5, alpha=1,
                                   nrep=100, ylimMax=0.2)

```

```{r, fig.path="./figureDRP/", dev=c("png", "pdf"), fig.width=9, fig.height=8}}
df_resultLassoPCA5 <- melt(resultLassoPCA5)
colnames(df_resultLassoPCA5) <- c("omic", "drug", "R2", "run")
summaryR25 <- df_resultLassoPCA5 %>% group_by(omic, drug) %>% 
  dplyr::summarise(meanR2=mean(R2),sdR2 = sd(R2), nR2 = length(R2)) %>%
  mutate(seR2 = sdR2/sqrt(nR2))

ggplot(summaryR25, aes(x=omic, y=meanR2, fill=omic, group=omic))+
    geom_bar(stat = "identity") +  scale_fill_manual(values=colMod) + 
    geom_errorbar(aes(ymax = meanR2 + sdR2,ymin = meanR2 - sdR2), position = "dodge", width = 0.25) +facet_wrap(~drug, ncol=4) +theme_bw(base_size = 18) +ylab(bquote(R^2)) + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.text = element_text(colour = "black")) +xlab("")+guides(fill=guide_legend(title="Data type")) 
```

```{r}
resultLassoPCA5 <-R2ForPenRegPlusViz(Z = ZPCsscaled, drugviab = drug, nfolds=3, alpha=1,
                                   nrep=100, ylimMax=0.3)
```


```{r}
X <- cbind(
  rna,
  prot,
  clin
)

stopifnot(!anyNA(X))
```

```{r}
fit <- cv.glmnet(
  x = X,
  y = y,
  alpha = 1,           # LASSO
  family = "gaussian",
  standardize = FALSE  # already scaled
)

R2 <- 1-min(fit$cvm)/fit$cvm[1]

R2
```

```{r, include=FALSE, eval=FALSE}
fit <- cv.glmnet(
  x = rna,
  y = y,
  alpha = 1,           # LASSO
  family = "gaussian",
  standardize = FALSE  # already scaled
)

R2 <- 1-min(fit$cvm)/fit$cvm[1]
R2
```





