---
title: "MOFA T-ALL"
author: "Shubham Agrawal"
date: "2026-06-19"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(tidyverse)
library(DESeq2)
library(MOFA2)
library(readxl)
source("helper.R")
```

# Load the data

```{r}
seProt <- readRDS("objects/protoemics.Rds")
dds <- readRDS("objects/ddsFiltered.Rds")
seDRP <- readRDS("objects/drpObject.Rds")
```

```{r}
snv <- readRDS("snvWES.rds")
cnv <- readRDS("cnvWES.rds")
```

# Load the metadata file

```{r}
metadataMofa <- read_csv("finalMetadata_Mofa.csv")
head(metadataMofa)
```
# Filter for T-ALL

```{r}
dim(metadataMofa)
metadataTall <- metadataMofa[metadataMofa$diseaseType == "T-ALL", ]
dim(metadataTall)
metadataTall$drugSample <- as.character(metadataTall$drugSample)
metadataTall$sample <- as.character(metadataTall$sample)
```
# Normalize or filter the omics assays

```{r}
dds <- DESeq2::estimateSizeFactors(dds)
vsd <- DESeq2::varianceStabilizingTransformation(dds) 
```

```{r}
dim(cnv)
# filter out low frequency events
event_freq <- rowSums(cnv)
cnv_binary_filt <- cnv[event_freq >= 3, ]
dim(cnv_binary_filt)
```
```{r}
# helper to align a matrix to common_samples given the sample_map mapping
alignView <- function(mat, from_ids, mapping_df, common_ids, colname_in_map) {
  # mat: original matrix (rows = features, cols = samples, colnames = sample IDs used in mat)
  # from_ids: character vector of sample IDs in mat (should equal colnames(mat))
  # mapping_df: the sample_map data.frame
  # common_ids: mapping_df$commonID
  # colname_in_map: column name in mapping_df that links to this view (e.g. "prot_sample")
  out <- matrix(NA_real_, nrow = nrow(mat), ncol = length(common_ids),
                dimnames = list(rownames(mat), as.character(common_ids)))
  # for each row in mapping, if sample exists in mat, copy column
  for (i in seq_along(common_ids)) {
    mapped_name <- mapping_df[[colname_in_map]][i]
    if (!is.na(mapped_name) && mapped_name %in% colnames(mat)) {
      out[, i] <- mat[, mapped_name]
    } else {
      # leave NA column
    }
  }
  return(out)
}

```

```{r}
commonSamples <- metadataTall$sample
```

```{r}
# WES SNV view 
snv_mat <- as.matrix(snv)  
snvView <- alignView(snv_mat, colnames(snv_mat), metadataTall, commonSamples, "wesSample")
```

```{r}
# WES CNV view 
cnv_mat <- as.matrix(cnv_binary_filt)  
cnvView <- alignView(cnv_mat, colnames(cnv_mat), metadataTall, commonSamples, "wesSample")
```



```{r}
# Proteomics view (assay might already be normalized, imputed)
protMat <- as.matrix(assay(seProt))  
# ensure colnames are the sample names used in sample_map$prot_sample
protView <- alignView(protMat, colnames(protMat), metadataTall, commonSamples, "protSample")

# Drug view (logAUC)
drugMat <- as.matrix(assay(seDRP))      # rows = drugs, cols = sampleIDs (drug sampleIDs)
drugView <- alignView(drugMat, colnames(drugMat), metadataTall, commonSamples, "drugSample")

# RNA counts view: use normalized counts assay from DESeq2 object
countsMat <- as.matrix(assay(vsd))
rnaView <- alignView(countsMat, colnames(countsMat), metadataTall, commonSamples, "rnaSample")
```
```{r}
# Optional: sanity check dims and names
dim(protView); dim(rnaView); dim(drugView); dim(snvView) ; dim(cnvView)
colnames(protView)  # should be common_samples (or same vector across views)
```
```{r}
protView1 <- protView[, 1:(ncol(protView) - 2)]
rnaView1 <- rnaView[, 1:(ncol(rnaView) - 2)]
drugView1 <- drugView[, 1:(ncol(drugView) - 2)]
snvView1 <- snvView[, 1:(ncol(snvView) - 2)]
cnvView1 <- cnvView[, 1:(ncol(cnvView) - 2)]
```



# Feature selection and filtering

```{r}
selectTopVar <- function(mat, topN = 2000) {
  varVec <- matrixStats::rowVars(mat, na.rm = TRUE)
  keep <- order(varVec, decreasing = TRUE)[seq_len(min(topN, sum(!is.na(varVec))))]
  mat[keep, , drop = FALSE]
}

rnaViewFilt <- selectTopVar(rnaView1, topN = 3000)     
protViewFilt <- selectTopVar(protView1, topN = 2000)
drugViewFilt <- drugView1   # usually small number of drugs
snvViewFilt <- snvView1
cnvViewFilt <- cnvView1 # already filtered
```

# Create the MOFA data list and MOFA object

```{r}
dataListX <- list(
  Proteomics = protView1,
  Transcriptomics = rnaView1,
  Drug = drugView1,
  SNV = snvView1,
  CNV = cnvView1
)

# Create MOFA object
mofaObjX <- create_mofa(dataListX)
plot_data_overview(mofaObjX) +
  theme(text = element_text(size = 20, face = 'bold'))
```

```{r}
dataList <- list(
  Proteomics = protViewFilt,
  Transcriptomics = rnaViewFilt,
  Drug = drugViewFilt,
  SNV = snvViewFilt,
  CNV = cnvViewFilt
)

# Create MOFA object
mofaObj <- create_mofa(dataList)
```
```{r}
plot_data_overview(mofaObj) +
  theme(text = element_text(size = 20, face = 'bold'))
```

# Set data / model / training options

```{r}
dataOpts   <- get_default_data_options(mofaObj)
modelOpts  <- get_default_model_options(mofaObj)
trainOpts  <- get_default_training_options(mofaObj)

modelOpts$num_factors <- 15   
trainOpts$maxiter <- 1000
trainOpts$convergence_mode <- "slow"
trainOpts$seed <- 42
```

Important: make sure the modelOpts$likelihoods order exactly matches the order 
of dataList.

# Prepare and run the MOFA model

```{r}
# Prepare the MOFA object for training
mofaPrepared <- prepare_mofa(mofaObj,
                             data_options = dataOpts,
                             model_options = modelOpts,
                             training_options = trainOpts)

# Run (this will create an HDF5 file and train the model)
trainedMofa <- run_mofa(mofaPrepared, outfile = "mofaFullTrainedTALL_19062026.hdf5",
                        use_basilisk = TRUE)
```

# Load Mofa model 

```{r}
trainedMofa <- load_model("mofaFullTrainedTALL_19062026.hdf5")
```

# Add metadata to the model



```{r}
samples_metadata(trainedMofa) <- metadataTall[1:(nrow(metadataTall) - 2), ]
```

# Correlation between factors

A good sanity check is to verify that the Factors are largely uncorrelated. In MOFA there are no orthogonality constraints such as in Principal Component Analysis, but if there is a lot of correlation between Factors this suggests a poor model fit. Reasons? Perhaps you used too many factors or perhaps the normalization is not adequate. [Link](https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/CLL.html#correlation-between-factors)

```{r}
plot_factor_cor(trainedMofa)
```

# Plot variance decomposition

This plot shows the percentage of variance explained by each factor across each 
data modality (and group, if provided). It summarises the sources of variation 
from a complex heterogeneous data set in a single figure.

```{r}
plot_variance_explained(trainedMofa, x="view", y="factor")
```

```{r}
plot_variance_explained(trainedMofa, plot_total = TRUE)[[2]]
```


```{r}
# Using Mofa function
correlate_factors_with_covariates(trainedMofa, 
  covariates = c("diagnosisType", "sex", "Age", "cytogenetics", 
                 "riskGroup", "predictionResponse"), 
  plot="log_pval"
)
```

```{r}
factors_df <- as.data.frame(get_factors(trainedMofa))
factors_df$sample = rownames(factors_df)
colnames(factors_df) <- sub("^group[0-9]+\\.", "", colnames(factors_df))
head(factors_df)
```

```{r}
meta <- samples_metadata(trainedMofa)
meta <- meta[, c("sample", "diagnosisType", "sex", "Age", "cytogenetics", 
                 "riskGroup", "predictionResponse")]
head(meta)
```
```{r}
resAssc <- testAssociation(tabX = factors_df, tabY = meta, joinID = "sample", 
                           correlation_method = "pearson", 
                           plot = TRUE, pCut = 0.01, ifFdr = FALSE, pMax = 12, 
                           onlySignificant = FALSE)

p <- resAssc$plot
p
```


```{r}
plot_factor(trainedMofa, 
  factors = 2, 
  color_by = "riskGroup",
  add_violin = TRUE,
  dodge = TRUE
)
```












