---
title: "Mofa analysis B-ALL"
author: "Shubham Agrawal"
date: "2025-09-15"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(tidyverse)
library(DESeq2)
library(MOFA2)
```

# Load the data and extract T-ALL samples only

```{r}
seProt <- readRDS("protoemics.Rds")
print("Proteomics data")
seProt
dds <- readRDS("ddsFiltered.Rds")
print("Transcriptomics data")
dds
seDRP <- readRDS("drpObject.Rds")
print("Drug Response Profile data")
seDRP
```

```{r}
seProtB <- seProt[, seProt$diseaseType == "B-ALL"]
print("Proteomics B-ALL data")
seProtB
dds <- dds[, !is.na(dds$diseaseType) & !is.na(dds$diagnosisType)]
ddsB <- dds[, dds$diseaseType == "B-ALL"]
print("Transcriptomics B-ALL data")
ddsB
seDrpB <- seDRP[, seDRP$diseaseType == "B-ALL"]
print("Drug Response Profile B-ALL data")
seDrpB
```

# Create unique identifier

Get the colData from dds and drp objects
```{r}
cdProt <- as.data.frame(colData(seProtB))
#head(cdProt)
cdDDS <- as.data.frame(colData(ddsB))
#head(cdDDS)
cdDDS <- select(cdDDS, samples, patID, Protein_Sample_ID)
cdDRP <- as.data.frame(colData(seDrpB))
#head(cdDRP)
cdDRP <- select(cdDRP, Labeling.proteomics, sample_id)

cdjoin <- left_join(cdProt, cdDDS, by = c("sample" = "Protein_Sample_ID"))
cdjoin <- left_join(cdjoin, cdDRP, by = c("sample" = "Labeling.proteomics"))
cdjoin <- cdjoin[!duplicated(cdjoin$sample), ]
```

```{r}
cdFinal <- cdjoin[!duplicated(cdjoin$sample), ]
dim(cdFinal)
names(cdFinal)[names(cdFinal) == "sample"] <- "protSample"
names(cdFinal)[names(cdFinal) == "samples"] <- "rnaSample"
names(cdFinal)[names(cdFinal) == "sample_id"] <- "drugSample"
```

```{r}
if (!"sample" %in% colnames(cdFinal)) cdFinal$sample <- as.character(seq_len(nrow(cdFinal)))
rownames(cdFinal) <- cdFinal$sample
commonSamples <- cdFinal$sample
```

# Normalize RNAseq data

The model can handle three types of data: continuous (modelled with a gaussian likelihood), small counts (modelled with a Poisson likelihood) and binary measurements (modelled with a bernoulli likelihood). Non-gaussian likelihoods give non-optimal results, we recommend the user to apply data transformations to obtain continuous measurements. For example, for count-based data such as RNA-seq or ATAC-seq we recommend size factor normalisation + variance stabilisation (i.e. a log transformation).

```{r}
ddsB <- DESeq2::estimateSizeFactors(ddsB)
vsdB <- DESeq2::varianceStabilizingTransformation(ddsB) 
```


# Create feature × sample matrices for each view (fill NAs for missing samples)

MOFA expects views as lists of matrices with rows as features and columns as 
samples (same columns across views). Therefore, create a column for every 
commonSamples, filling NA for samples missing in a view.

```{r}
# helper to align a matrix to common_samples given the sample_map mapping
alignView <- function(mat, from_ids, mapping_df, common_ids, colname_in_map) {
  # mat: original matrix (rows = features, cols = samples, colnames = sample IDs used in mat)
  # from_ids: character vector of sample IDs in mat (should equal colnames(mat))
  # mapping_df: the sample_map data.frame
  # common_ids: mapping_df$commonID
  # colname_in_map: column name in mapping_df that links to this view (e.g. "prot_sample")
  out <- matrix(NA_real_, nrow = nrow(mat), ncol = length(common_ids),
                dimnames = list(rownames(mat), as.character(common_ids)))
  # for each row in mapping, if sample exists in mat, copy column
  for (i in seq_along(common_ids)) {
    mapped_name <- mapping_df[[colname_in_map]][i]
    if (!is.na(mapped_name) && mapped_name %in% colnames(mat)) {
      out[, i] <- mat[, mapped_name]
    } else {
      # leave NA column
    }
  }
  return(out)
}

# Proteomics view (assay might already be normalized, imputed)
protMat <- as.matrix(assay(seProtB))  
# ensure colnames are the sample names used in sample_map$prot_sample
protView <- alignView(protMat, colnames(protMat), cdFinal, commonSamples, "protSample")

# Drug view (logAUC)
drugMat <- as.matrix(assay(seDrpB))      # rows = drugs, cols = sampleIDs (drug sampleIDs)
drugView <- alignView(drugMat, colnames(drugMat), cdFinal, commonSamples, "drugSample")

# RNA counts view: use normalized counts assay from DESeq2 object
countsMat <- as.matrix(assay(vsdB))
rnaView <- alignView(countsMat, colnames(countsMat), cdFinal, commonSamples, "rnaSample")

# Optional: sanity check dims and names
dim(protView); dim(rnaView); dim(drugView)
colnames(protView)  # should be common_samples (or same vector across views)
```

# Feature selection and filtering

```{r}
selectTopVar <- function(mat, topN = 2000) {
  varVec <- matrixStats::rowVars(mat, na.rm = TRUE)
  keep <- order(varVec, decreasing = TRUE)[seq_len(min(topN, sum(!is.na(varVec))))]
  mat[keep, , drop = FALSE]
}

rnaViewFilt <- selectTopVar(rnaView, topN = 3000)     
protViewFilt <- selectTopVar(protView, topN = 2000)
drugViewFilt <- drugView   # usually small number of drugs
```

# Create the MOFA data list and MOFA object

```{r}
dataList <- list(
  Proteomics = protViewFilt,
  Transcriptomics = rnaViewFilt,
  Drug = drugViewFilt
)

# Create MOFA object
mofaObj <- create_mofa(dataList)
```
```{r}
plot_data_overview(mofaObj) +
  theme(text = element_text(size = 20, face = 'bold'))
```

# Set data / model / training options

```{r}
dataOpts   <- get_default_data_options(mofaObj)
modelOpts  <- get_default_model_options(mofaObj)
trainOpts  <- get_default_training_options(mofaObj)

modelOpts$num_factors <- 10   
trainOpts$maxiter <- 1000
```

# Prepare and run the MOFA model

```{r, message=FALSE}
# Prepare the MOFA object for training
mofaPrepared <- prepare_mofa(mofaObj,
                             data_options = dataOpts,
                             model_options = modelOpts,
                             training_options = trainOpts)

# Run (this will create an HDF5 file and train the model)
trainedMofa <- run_mofa(mofaPrepared, outfile = "mofaTrainedBALL.hdf5",
                        use_basilisk = TRUE)
```

# Add metadata to the model

```{r}
cdFinalsub <- subset(cdFinal, select =  -c(diseaseType))
samples_metadata(trainedMofa) <- cdFinalsub
```

# Visualizations

```{r}
plot_variance_explained(trainedMofa, x="view", y="factor")
```
```{r}
plot_variance_explained(trainedMofa, plot_total = TRUE)[[2]]
```
```{r}
plot_factor_cor(trainedMofa)
```
```{r}
plot_factors(trainedMofa, 
  factors = 1:5,
  color_by = "diagnosisType"
)
```

```{r}
correlate_factors_with_covariates(trainedMofa, 
  covariates = c("diagnosisType", "Sex", "Age", "cytogenetics", 
                 "riskGroup", "predictionResponse"), 
  plot="log_pval"
)
```
```{r}
plot_top_weights(trainedMofa,
 view = "Transcriptomics",
 factor = 1,
 nfeatures = 20,     # Top number of features to highlight
 scale = T           # Scale weights from -1 to 1
)
```
```{r}
plot_factor(trainedMofa, 
  factors = 1, 
  color_by = "diagnosisType",
  add_violin = TRUE,
  dodge = TRUE
)
```

